---
title: "DANA4820_Project_bernard"
author: "Bernard Bernabeo"
date: "2025-03-25"
output:
  word_document: default
  html_document: default
---

STUDY OBJECTIVE: How do immunization rates, safe water access, and air pollution levels impact infant mortality across countries over a 10-year period?



```{r}
#Load libraries
library(dplyr)
library(ggplot2)
library(MASS)
library(countrycode)
library(car) 
library(tidyr)

```

```{r}
#Load dataset
global_health = read.csv("global_health.csv")  

```

```{r}
#Data structure
str(global_health)
summary(global_health)
```

```{r}
# Add continent column based on country names
global_health$Continent = countrycode(global_health$Country, origin = "country.name", 
                  destination = "continent")

# View a few rows to confirm
head(global_health)

# View dataset 
View(global_health)
```

```{r}
# Convert Year to factor (categorical)
global_health$Year <- as.factor(global_health$Year)
```



```{r}
#Total missing values
sum(is.na(global_health))

colSums(is.na(global_health))
rowSums(is.na(global_health))
```

```{r}
# Check distributions using histograms (raw data)
par(mfrow = c(2,2))  # Arrange plots in a 2x2 grid
hist(global_health$Immunization_Rate, main = "Immunization Rate", col = "lightblue", breaks = 30)
hist(global_health$Safe_Water_Access_Percent, main = "Safe Water Access Percent", col = "lightgreen", breaks = 30)
hist(global_health$Air_Pollution, main = "Air Pollution", col = "lightcoral", breaks = 30)
hist(global_health$Total_Population, main = "Total Population", col = "lightgray", breaks = 30)
```


```{r}
library(dplyr)
library(tidyr)

# Median imputation for missing values 
data_filled <- global_health %>% 
  dplyr::select(Country, GDP_Per_Capita, Safe_Water_Access_Percent, Year, Immunization_Rate, Air_Pollution, Infant_Deaths, Fertility_Rate) %>%
  group_by(Year) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))%>%
ungroup()
```

```{r}
# Summary statistics after imputation
summary(data_filled)
```
```{r}
df_complete <- data_filled[complete.cases(data_filled), ]
df_complete$Year <- as.factor(df_complete$Year)
View(df_complete)
```

```{r}
summary(df_complete)
```



```{r}
# Check distributions using histograms (raw data)
par(mfrow = c(2,2))  # Arrange plots in a 2x2 grid
hist(df_complete$Immunization_Rate, main = "Immunization Rate", col = "lightblue", breaks = 30)
hist(df_complete$Safe_Water_Access_Percent, main = "Safe Water Access Percent", col = "lightgreen", breaks = 30)
hist(df_complete$Air_Pollution, main = "Air Pollution", col = "lightcoral", breaks = 30)
#hist(data_filled$Total_Population, main = , col = "lightgray", breaks = 30)
```



```{r}
# Boxplots to detect outliers
boxplot(df_complete$Immunization_Rate, main = "Immunization Rate", col = "lightblue")
boxplot(df_complete$Safe_Water_Access_Percent, main = "Safe Water Access Percent", col = "lightgreen")
boxplot(df_complete$Air_Pollution, main = "Air Pollution", col = "lightcoral")
#boxplot(global_health$Total_Population, main = "Total Population", col = "lightgray")
```


```{r}
# Fit Poisson regression model (Infant Mortality as the response variable)
model <- glm(Infant_Deaths ~ GDP_Per_Capita + Safe_Water_Access_Percent  + Immunization_Rate + Air_Pollution + Infant_Deaths + Year + Fertility_Rate, family = poisson,
                   data = df_complete)
summary(model)
```
```{r}
# Check for overdispersion
dispersion_model <- sum(residuals(model, type = "pearson")^2) / model$df.residual
dispersion_model  # If >1.5, consider Negative Binomial regression

```
```{r}
df_complete$Year <- as.numeric(as.character(df_complete$Year))
```


```{r}
cor(df_complete[, c("GDP_Per_Capita", "Safe_Water_Access_Percent", "Immunization_Rate", "Air_Pollution", "Infant_Deaths", "Year" , "Fertility_Rate")], use = "pairwise.complete.obs")
```

```{r}
# Negative Binomial regression model (Infant Mortality as the response variable)
library(MASS)
nb_model <- glm.nb(Infant_Deaths ~ GDP_Per_Capita + Safe_Water_Access_Percent + Immunization_Rate + Air_Pollution + Year, 
                   data = df_complete,
                   start = rep(0.1, 6))  # Adjust the number of starting values based on predictors
summary(nb_model)
```

```{r}
poisson_model <- glm(Infant_Deaths ~ GDP_Per_Capita + Safe_Water_Access_Percent + Immunization_Rate + Air_Pollution + Year, 
                     data = df_complete, 
                     family = poisson)
summary(poisson_model)
```


#### predicted logit [pi (Infant Deaths)] = beta0 + beta1 * Immunization_Rate + beta2 * Safe_Water_Access_Percent + beta3 * Air_Pollution + beta4 * Total_Population + beta5 * Year

#### predicted logit [pi (Infant Deaths)] = 8.464 + (-0.252) * Immunization_Rate + (-0.967) * Safe_Water_Access_Percent + (0.604) * Air_Pollution + (2.39) * Total_Population + (0.044) * Year


H0: betai = 0 Predictor/s has no contribution on the prediction of Infant Deaths.
Ha: betai != 0 At least one of the predictors has effect on Infant Deaths.

```{r}
# Converting log-transformed coefficients from Negative Binomial model into probabilities
exp(coef(nb_model))
```
```{r}
# Variance Inflation Factor (VIF) for multicollinearity check
vif(nb_model)
```


#### INTERPRETATION:
Intercept(1.649e-10):This is the baseline value for the infant death rate when all predictors are set to zero. Since in real life, these predictors can't be zero, it is practically meaningful to interpret its value. 
Immunization Rate(0.7773): A 1-unit increase in immunization rate reduces infant deaths by 22.3%
Safe Water Access Percent(0.3801): A 1% increase in safe water access reduces infant deaths by 62%.
Air Pollution(1.8289): A 1-unit increase in air pollution increase infant death by 83%.
Total Population(10.9140): A 1-unit increase in population multiplies infant deaths by 10.91 times
Year(1.045): Since p-value (0.261) is > than alpha (0.05 or 0.1), we fail to reject H0.Year does not appear to contribute meaningfully to explaining variation in infant deaths. 


```{r}
# Check for overdispersion
dispersion <- sum(residuals(nb_model, type = "pearson")^2) / nb_model$df.residual
dispersion  # If >1.5, consider Negative Binomial regression

```
#Categorizing Air pollution 
```{r}
global_health_scaled$Air_Pollution_Category <- cut(global_health_scaled$Air_Pollution, 
                                   breaks = c(-Inf, 50, 100, Inf),  
                                   labels = c("Good", "Moderate", "Unhealthy"))
View(global_health_scaled)
```

