---
title: "Project-Objective 2"
author: "Group 1"
date: "2025-03-26"
output:
  word_document: default
---
# What economic and social factors most influence lower life expectancy in gender?


## Dataset

```{r}
# Load necessary libraries
library(dplyr)  # Data manipulation
library(ggplot2)  # Data visualization
library(caret)  # Machine learning and feature selection
library(corrplot)  # Correlation plot
library(car)  # For regression diagnostics
library(countrycode)
```



```{r}
# Load dataset
global_health <- read.csv("global_health.csv")

# Inspect dataset
str(global_health)  # Check structure
summary(global_health)  # Get basic statistics
head(global_health)  # Preview first few rows
```


```{r}
# Add continent column based on country names
global_health$Continent = countrycode(global_health$Country, origin = "country.name", 
                  destination = "continent")

# View a few rows to confirm
head(global_health)

# View dataset 
View(global_health)
```


```{r}
# Convert Year to factor (categorical)
global_health$Year <- as.factor(global_health$Year)
```


```{r}
#Total missing values
sum(is.na(global_health))

colSums(is.na(global_health))

```


```{r}
# Check distributions using histograms (raw data)
par(mfrow = c(3,2))  # Arrange plots in a 2x2 grid
hist(global_health$Life_Expectancy_Male, main = "Life Expectancy Male", col = "lightblue", breaks = 30)
hist(global_health$Life_Expectancy_Female, main = "Life Expectancy Female", col = "lightgreen", breaks = 30)
hist(global_health$GDP_Per_Capita, main = "GDP_Per_Capita", col = "lightcoral", breaks = 30)
hist(global_health$Urban_Population_Percent, main = "Urban_Population_Percent", col = "lightgray", breaks = 30)
hist(global_health$Unemployment_Rate, main = "Unemployment_Rate", col = "lightpink", breaks = 30)
hist(global_health$Safe_Water_Access_Percent, main = "Safe_Water_Access_Percent", col = "lightyellow", breaks = 30)
```


```{r}
library(dplyr)
library(tidyr)
data_filled <- global_health %>%
  select(Country, Life_Expectancy, Life_Expectancy_Female, Life_Expectancy_Male, GDP_Per_Capita, Unemployment_Rate, Safe_Water_Access_Percent, Urban_Population_Percent) %>%
  group_by(Country) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()
```
```{r}
summary(data_filled)
```



```{r}
global_health <- data_filled[complete.cases(data_filled), ]

```


```{r}
# Summary statistics after imputation
summary(global_health)
```

```{r}
# Check distributions using histograms (raw data)
par(mfrow = c(3,2))  # Arrange plots in a 2x2 grid
hist(global_health$Life_Expectancy_Male, main = "Life Expectancy Male", col = "lightblue", breaks = 30)
hist(global_health$Life_Expectancy_Female, main = "Life Expectancy Female", col = "lightgreen", breaks = 30)
hist(global_health$GDP_Per_Capita, main = "GDP_Per_Capita", col = "lightcoral", breaks = 30)
hist(global_health$Urban_Population_Percent, main = "Urban_Population_Percent", col = "lightgray", breaks = 30)
hist(global_health$Unemployment_Rate, main = "Unemployment_Rate", col = "lightpink", breaks = 30)
hist(global_health$Safe_Water_Access_Percent, main = "Safe_Water_Access_Percent", col = "lightyellow", breaks = 30)
```






```{r}
# Boxplots to detect outliers
boxplot(global_health$Life_Expectancy_Male, main = "Life_Expectancy_Male", col = "lightblue")
boxplot(global_health$Life_Expectancy_Female, main = "Life_Expectancy_Female", col = "lightgreen")
boxplot(global_health$GDP_Per_Capita, main = "GDP_Per_Capita", col = "lightcoral")
boxplot(global_health$Urban_Population_Percent, main = "Urban_Population_Percent", col = "lightgray")
boxplot(global_health$Unemployment_Rate, main = "Unemployment_Rate", col = "lightpink")
boxplot(global_health$Safe_Water_Access_Percent, main = "Safe_Water_Access_Percent", col = "lightyellow")
```
```{r}
# Combined boxplot for Male and Female Life Expectancy
boxplot(global_health$Life_Expectancy_Male,
        global_health$Life_Expectancy_Female,
        names = c("Male", "Female"),
        main = "Life Expectancy by Gender",
        col = c("#88BBD6", "#003366"),
        ylab = "Life Expectancy (Years)")

```



```{r}
# Categorías: Low (≤75), High (>75)
global_health$Life_Expectancy_Category_Male <- cut(global_health$Life_Expectancy_Male,
                                            breaks = c(-Inf, 75, Inf),
                                            labels = c("Low", "High"))

global_health$Life_Expectancy_Category_Female <- cut(global_health$Life_Expectancy_Female,
                                              breaks = c(-Inf, 75, Inf),
                                              labels = c("Low", "High"))

```



```{r}
# GDP per Capita (Low / High based on median)
global_health$GDP_Category <- ifelse(global_health$GDP_Per_Capita < median(global_health$GDP_Per_Capita, na.rm = TRUE),
                              "Low GDP", "High GDP")

# Urban Population
global_health$Urban_Category <- ifelse(global_health$Urban_Population_Percent < median(global_health$Urban_Population_Percent, na.rm = TRUE),
                                "Low Urban", "High Urban")

# Unemployment Rate
global_health$Unemployment_Category <- ifelse(global_health$Unemployment_Rate < median(global_health$Unemployment_Rate, na.rm = TRUE),
                                   "Low Unemployment Rate", "High Unemployment Rate")

# Safe_Water_Access_Percent
global_health$Safe_Water_Access_Category <- ifelse(global_health$Safe_Water_Access_Percent < median(global_health$Safe_Water_Access_Percent, na.rm = TRUE),
                                   "Low Safe Water Access", "High Safe Water Access")
```


# Female
## GDP Category vs. Female Life Expectancy

```{r}
# Create contingency table
table_female_gdp <- table(global_health$Life_Expectancy_Category_Female,
                          global_health$GDP_Category)
table_female_gdp

# Perform Chi-Square Test
chisq.test(table_female_gdp, correct = FALSE)

```
### Interpretation

-Since the p-value is far less than 0.05, we reject the null hypothesis.
-This means there is a statistically significant association between GDP category and female life expectancy.
-Countries with higher GDP are more likely to have higher female life expectancy.

## Urban Population Category vs. Female Life Expectancy

```{r}
table_female_urban <- table(global_health$Life_Expectancy_Category_Female,
                            global_health$Urban_Category)
table_female_urban

chisq.test(table_female_urban,correct = FALSE)

```
### Interpretation

-Since p < 0.05, you reject the null hypothesis.
-There is a statistically significant relationship between urban population level and female life expectancy.
-Countries with higher urbanization tend to have higher female life expectancy.

## Unemployment Category vs. Female Life Expectancy

```{r}
table_female_unemp <- table(global_health$Life_Expectancy_Category_Female,
                            global_health$Unemployment_Category)
table_female_unemp

chisq.test(table_female_unemp, correct = FALSE)
```
### Interpretion

-p < 0.05, so you reject the null hypothesis.
-There is a significant association between unemployment rate and female life expectancy.
-Even among high unemployment countries, many still have high life expectancy, but the association remains statistically significant.


## Safe Water Access Category vs. Female Life Expectancy

```{r}
table_female_water <- table(global_health$Life_Expectancy_Category_Female,
                            global_health$Safe_Water_Access_Category)
table_female_water

chisq.test(table_female_water, correct = FALSE)

```
### Interpretation

-There is a statistically significant association between safe water access and female life expectancy (p < 0.05).
-This means that countries with high access to safe water tend to have higher female life expectancy.
-In countries where people have better access to safe water, women are more likely to live longer.


# Male
## GDP Category vs. Male Life Expectancy

```{r}
# Create contingency table
table_Male_gdp <- table(global_health$Life_Expectancy_Category_Male,
                          global_health$GDP_Category)
table_Male_gdp

# Perform Chi-Square Test
chisq.test(table_Male_gdp, correct = FALSE)

```
### Interpretation

-There is a statistically significant association between GDP per capita and male life expectancy.
-Countries with higher GDP are far more likely to have higher male life expectancy.
-In contrast, low GDP countries are overwhelmingly concentrated in the low life expectancy group.

## Urban Population Category vs. Male Life Expectancy

```{r}
table_Male_urban <- table(global_health$Life_Expectancy_Category_Male,
                            global_health$Urban_Category)
table_Male_urban

chisq.test(table_Male_urban, correct = FALSE)

```
### Interpretation

-There is a statistically significant association between urban population and male life expectancy.
-Countries with high urbanization are much more likely to have higher male life expectancy.
-On the other hand, low urbanization is strongly associated with low life expectancy in males.

## Unemployment Category vs. Male Life Expectancy

```{r}
table_Male_unemp <- table(global_health$Life_Expectancy_Category_Male,
                            global_health$Unemployment_Category)
table_Male_unemp

chisq.test(table_Male_unemp, correct = FALSE)
```
### Interpretation 
-There is no statistically significant association between unemployment rate and male life expectancy.
-The p-value (0.7682) is much greater than 0.05, indicating that male life expectancy appears to be independent of unemployment levels in this dataset.

## Safe Water Access Category vs. Male Life Expectancy

```{r}
table_Male_water <- table(global_health$Life_Expectancy_Category_Male,
                            global_health$Safe_Water_Access_Category)
table_Male_water

chisq.test(table_Male_water, correct = FALSE)

```
### Interpretation

-There is a statistically significant association between safe water access and male life expectancy.
-Countries with high access to safe water are much more likely to have higher male life expectancy.
-Countries with poor water access tend to have lower male life expectancy.


## Visualizations

```{r}
library(ggplot2)

```

```{r}
# Combine male and female data for faceted plotting
library(dplyr)

data_female <- global_health %>%
  select(GDP_Category, Urban_Category, Unemployment_Category, Safe_Water_Access_Category,
         Life_Expectancy_Category_Female) %>%
  mutate(Gender = "Female",
         Life_Expectancy_Category = Life_Expectancy_Category_Female)

data_male <- global_health %>%
  select(GDP_Category, Urban_Category, Unemployment_Category, Safe_Water_Access_Category,
         Life_Expectancy_Category_Male) %>%
  mutate(Gender = "Male",
         Life_Expectancy_Category = Life_Expectancy_Category_Male)

# Combine into one dataset
gender_data <- bind_rows(data_female, data_male)

```

## Comparative Plot – Life Expectancy by GDP and Gender


```{r}
ggplot(gender_data, aes(x = GDP_Category, fill = Life_Expectancy_Category)) +
  geom_bar(position = "fill") +
  facet_wrap(~Gender) +
  labs(title = "Life Expectancy by GDP and Gender",
       x = "GDP Category",
       y = "Proportion",
       fill = "Life Expectancy") +
  scale_fill_manual(values = c("Low" = "#88BBD6", "High" = "#003366")) +
  theme_minimal()
```

## Comparative Plot – Life Expectancy by Urbanization and Gender

```{r}
ggplot(gender_data, aes(x = Urban_Category, fill = Life_Expectancy_Category)) +
  geom_bar(position = "fill") +
  facet_wrap(~Gender) +
  labs(title = "Life Expectancy by Urbanization and Gender",
       x = "Urbanization Level",
       y = "Proportion",
       fill = "Life Expectancy") +
  scale_fill_manual(values = c("Low" = "#88BBD6",
                             "High" = "#003366"))  +
  theme_minimal()

```


## Comparative Plot – Life Expectancy by Unemployment and Gender

```{r}

gender_data$Unemployment_Category <- ifelse(
  gender_data$Unemployment_Category == "High Unemployment Rate", 
  "High\nUnemployment\nRate", 
  "Low\nUnemployment\nRate"
)

ggplot(gender_data, aes(x = Unemployment_Category, fill = Life_Expectancy_Category)) +
  geom_bar(position = "fill") +
  facet_wrap(~Gender) +
  labs(title = "Life Expectancy by Unemployment and Gender",
       x = "Unemployment Level",
       y = "Proportion",
       fill = "Life Expectancy") +
  scale_fill_manual(values = c("Low" = "#88BBD6", "High" = "#003366")) +
  theme_minimal()

```


## Comparative Plot – Life Expectancy by Safe Water Access and Gender

```{r}
gender_data$Safe_Water_Access_Category <- ifelse(
  gender_data$Safe_Water_Access_Category == "High Safe Water Access",
  "High\nSafe Water\nAccess",
  "Low\nSafe Water\nAccess"
)

ggplot(gender_data, aes(x = Safe_Water_Access_Category, fill = Life_Expectancy_Category)) +
  geom_bar(position = "fill") +
  facet_wrap(~Gender) +
  labs(title = "Life Expectancy by Safe Water Access and Gender",
       x = "Safe Water Access",
       y = "Proportion",
       fill = "Life Expectancy") +
  scale_fill_manual(values = c("Low" = "#88BBD6", "High" = "#003366"))+
  theme_minimal()

```

# GLM

```{r}
model_female <- glm(Life_Expectancy_Category_Female ~ GDP_Category + Urban_Category +
             Safe_Water_Access_Category + Unemployment_Category,
             family = binomial(link="logit"), data = global_health)

summary(model_female)


```

```{r}
exp(coef(model_female))

```
### Interpretation

-Holding all other variables constant, the odds of having high female life expectancy in countries with low GDP are approximately 77% lower than in countries with high GDP. Strong evidence of a GDP effect (p < 0.001).  
-Countries with low urbanization have about 90% lower odds of high female life expectancy compared to highly urbanized countries, after controlling for other variables. Strong urbanization effect (p < 0.001).  
-The odds of high female life expectancy in countries with poor access to safe water are about 91% lower than those with good access, controlling for GDP, urbanization, and unemployment. Very strong effect (p < 0.001).  
-Unemployment effect is not statistically significant (p = 0.139).  


```{r}
model_male <- glm(Life_Expectancy_Category_Male ~ GDP_Category + Urban_Category +
             Safe_Water_Access_Category + Unemployment_Category,
             family = binomial(link="logit"), data = global_health)

summary(model_male)
```


```{r}
exp(coef(model_male))

```
### Interpretation

-Holding all other variables constant, countries with low GDP have 93% lower odds of high male life expectancy compared to those with high GDP.  Strong GDP effect (p < 0.001).  
-Countries with low urbanization have 85% lower odds of high male life expectancy than highly urbanized countries, adjusting for other variables. Strong urbanization effect (p < 0.001).  
-Countries with poor access to safe water have 91% lower odds of high male life expectancy than countries with good access. Very strong effect (p < 0.001).  
-Countries with low unemployment have more than twice the odds (OR ≈ 2.16) of high male life expectancy compared to those with high unemployment. This time, the effect is statistically significant (p < 0.001).
